---
title: "Plamid cluster distribution"
author: "Sneha Sundar"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Script/Statistics for Fig S6.

Is F-like transfer operon completeness associated to specific plasmid clusters? 

We used MOB-recon from MOB-Suite (v.3.1.9, default parameters, using a set of
closed Enterobacteriaceae genomes for chromosome depletion) for plasmid recon-
struction and clustering. The default mash distance threshold of 0.05 was used to
assign plasmid cluster identities. Each genome is represented as a binary vector that
describes the presence and absence of these plasmid cluster IDs.
The plasmid clusters specifically associated with the F-like plasmid-associated contigs identified using our transfer gene finding protocol were considered to be F-like plasmid
clusters. 

```{r}
## Libraries

library(here)
library(tidyverse)

source(here("code/000.tools/helpers.R"))
source(here("code/000.tools/plot.R"))
```

## Load main dataset

Input summary table for all genomes. 
```{r}
genome_summary_df <- read.table(here("data/processed/002.find_genes/pires1200_onlyblastcontigs_rmnonF-like_trantotraf/genome_tra_count_summary.tsv"),header = T,sep = '\t', quote = "")


genome_summary_df
```


Plots folder
```{r}
plots_folder <- here("data/processed/manuscript_plots/006.plasmid.context")

```

## Bakta contig-level summary 


```{r}
INPUT_BAKTA_FILE <- here("data/processed/002.find_genes/pires1200_onlyblastcontigs_rmnonF-like_trantotraf/pires1200bakta_onlyblastcontigs_rmnonF-like_trantotraf.tsv")

pires1200_bakta <- load_processed_bakta(INPUT_BAKTA_FILE) 

pires1200_bakta_contig_summary <- pires1200_bakta %>% group_by(genome_contig) %>% summarise(n.genes = n()) %>% 
  mutate(genome = str_split_i(genome_contig,"_",1))

```



## Results from reconstructing plasmid sequences using mob_recon


```{r}
mob_recon_results <- read.table(here("data/processed/002.find_genes/pires1200/ecoli/mob_recon/all.mge_report.txt"),header = T,sep = '\t', quote = "")


mob_recon_results <- mob_recon_results %>% 
  mutate(contig_id = str_split_i(contig_id," ",1)) %>% 
  mutate(genome_contig = paste0(sample_id,"_",contig_id),.after = 1)


mob_typer_results <- read.table(here("data/processed/002.find_genes/pires1200/ecoli/mob_recon/all.mobtyper_results.txt"),header = T,sep = '\t', quote = "")


mob_typer_results <- mob_typer_results %>% mutate(genome = str_split_i(sample_id,":",1),.before = 2)

```


## What plasmid clusters (if any) are the contigs we identify as F-like assigned to? 


```{r}
mob_recon_results_F_like  <- pires1200_bakta_contig_summary %>% left_join(mob_recon_results) %>% group_by(genome) %>% summarise(F_like_plasmid_cluster = paste(unique(primary_cluster_id),collapse = "_")) %>% mutate(F_like_plasmid_cluster = str_remove(F_like_plasmid_cluster,"_NA|NA_")) %>% 
  mutate(F_like_plasmid_cluster = if_else(F_like_plasmid_cluster=="NA"|F_like_plasmid_cluster=="",NA,F_like_plasmid_cluster))


mob_recon_results_F_like %>% group_by(F_like_plasmid_cluster) %>% summarise(n())


genome_summary_df <- genome_summary_df %>% left_join(mob_recon_results_F_like, join_by(target_genome == genome))

```


## Proportion of genomes in which the transfer operon is present (>= 5 transfer genes) that were assigned a plasmid cluster
```{r}
x <- genome_summary_df %>% filter(bakta.n_distinct_hits >=5) %>% pull(F_like_plasmid_cluster) %>% table(useNA = "ifany") %>% prop.table()


1 - x[length(x)]
```


## What about genomes classified as missing F-like transfer operons? Only two have been assigned a plasmid cluster. 

```{r}
genome_summary_df %>% filter(bakta.n_distinct_hits <5) %>% pull(F_like_plasmid_cluster) %>% table(useNA = "ifany") 

```

## Figure: Tra operon functional completeness distribution of F-like plasmid clusters

```{r}
p <- genome_summary_df %>% filter(!is.na(F_like_plasmid_cluster)) %>% ggplot() + geom_bar(aes(y=F_like_plasmid_cluster,fill=functional_category)) + 
  facet_wrap(vars(Generic_Host)) + 
  scale_fill_manual(values=functional_colours,name="Functional completeness") + 
  ylab("F-like plasmid cluster ID") +
  theme_manuscript 

p

ggsave(paste0(plots_folder,"/","plasmid_clusters_host_functional_completeness.pdf"),p, units = "cm",height = 20, width = 24 , dpi = 600)
```

