---
title: "Contig-level distribution of transfer genes/gene synteny"
author: "Sneha Sundar"
date: "`r Sys.Date()`"
output: html_document
---

Script and statistics for Fig S4


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(here)
library(tidyverse)
source(here("code/000.tools/helpers.R"))
source(here("code/000.tools/plot.R"))



plots_folder <- here("data/processed/manuscript_plots/003.count")



```

# Contig-level distribition of transfer gene hits

## Load required data

```{r}
#bakta annotation file
INPUT_BAKTA_FILE <- here("data/processed/002.find_genes/pires1200_onlyblastcontigs_rmnonF-like_trantotraf/pires1200bakta_onlyblastcontigs_rmnonF-like_trantotraf.tsv")

pires1200_bakta <- load_processed_bakta(INPUT_BAKTA_FILE) 

#genome level summary
genome_summary_df <- read.table(here("data/processed/002.find_genes/pires1200_onlyblastcontigs_rmnonF-like_trantotraf/genome_tra_count_summary.tsv"),header = T,sep = '\t', quote = "")

```


## Figure: In how many contigs are transfer gene hits found? 

```{r}
p1 <- genome_summary_df %>% ggplot() +
geom_histogram(aes(x=n.contigs.bakta),bins = 30) + 
  ylab("Number of genomes") +
 # scale_x_continuous(limits = c(0,30))+
  xlab("Number of contigs with detected transfer gene hits")+
  theme_manuscript

p1
ggsave(paste0(plots_folder,"/","bakta_hist_contigs_with_hits.svg"),p1, units = "cm",height = 8, width = 16)
```


What proportion of transfer gene hits are found across multiple contigs?

```{r}
(genome_summary_df %>% filter(n.contigs.bakta>1) |> nrow())/1200
```
```{r}
(genome_summary_df %>% filter(n.contigs.bakta>2) |> nrow())/1200

```

## Split across tra operon completeness category

```{r}

p2 <- genome_summary_df %>% ggplot() + geom_histogram(aes(x=n.contigs.bakta,fill=functional_category), bins = 25) + facet_wrap(vars(functional_category), ncol=3) + 
  xlab("Number of contigs with detected transfer gene hits")+
  ylab("Number of genomes")+
  theme_manuscript + 
  scale_fill_manual(values = functional_colours, name = "Functional completeness") + 
  theme(legend.position = "top")

p2
ggsave(paste0(plots_folder,"/","bakta_hist_contigs_with_hits_by_functional_category.svg"),p2, units = "cm",height = 8, width = 16, dpi = 600)

```

## Average number of contigs on which transfer genes are found for genomes with complete and incomplete transfer operons

```{r}
complete <- genome_summary_df %>% filter(functional_category =="complete") %>% pull(n.contigs.bakta)

incomplete <- genome_summary_df %>% filter(functional_category =="incomplete") %>% pull(n.contigs.bakta)


mean(complete)
mean(incomplete)
```


```{r}
wilcox.test(n.contigs.bakta~functional_category, data = genome_summary_df %>% filter(functional_category %in% c("complete","incomplete")) %>% mutate(functional_category=as.factor(functional_category)),alternative = "less")


pairwise.wilcox.test(genome_summary_df$n.contigs.bakta,genome_summary_df$functional_category)
```


# Gene synteny


## Get gene orders of all contigs contining F-like transfer genes and check if they are in the same order found on the classic F plasimd. 
```{r}

gene_orders <- sapply(unique(pires1200_bakta$genome_contig), extract_tra_gene_order, pires1200_bakta)


positional_orders <- sapply(gene_orders, get_positional_order)

is_F_order_vec <- sapply(unique(pires1200_bakta$genome_contig), check_if_F_order, pires1200_bakta)


pires1200_bakta_contig_summary <- pires1200_bakta %>% group_by(genome_contig) %>% summarise(n.genes = n()) %>% 
  mutate(is_F_order = is_F_order_vec) 



pires1200_bakta_contig_summary <- pires1200_bakta_contig_summary %>% 
  mutate(genome = str_split_i(pires1200_bakta_contig_summary$genome_contig,"_",1))



pires1200_bakta_summary.Forder <- pires1200_bakta_contig_summary %>% group_by(genome) %>% summarise(prop.contigs.F_order = sum(is_F_order,na.rm=T)/sum(!is.na(is_F_order)), n.contigs = n(),n.contigs.more1gene =sum(!is.na(is_F_order)))

pires1200_bakta_summary.Forder

pires1200_bakta_summary.Forder <- pires1200_bakta_summary.Forder %>% mutate(all.F_order = if_else(prop.contigs.F_order == 1,true="yes",false = "no",missing="only 1 transfer gene on contig"))


```


# Figure: Is F order gene synteny conserved even when hits are distributed across multiple contigs? 

```{r}


p <- pires1200_bakta_summary.Forder %>% 
  mutate(all.F_order = factor(all.F_order,levels = c("yes","no","only 1 transfer gene on contig"))) %>% 
  ggplot() + 
  geom_histogram(aes(x=n.contigs,fill=all.F_order),bins = 25) + 
  xlab("Number of contigs with detected transfer gene hits") + 
  ylab("Number of genomes") + 
  scale_fill_manual(values = ternary_colours,name = "Is F plasmid gene order\nconserved?") +
  themeBW_manuscript

p

ggsave(paste0(plots_folder,"/","bakta_gene_synteny_distribution.pdf"),p, units = "cm",height = 8, width = 16, dpi = 600)

```

## What proportion of genomes where gene synteny can be assessed (have at least one contig with more than 1 transfer gene) have a conserved F plasmid gene synteny? 

```{r}
genome_summary_df <- genome_summary_df %>% left_join(pires1200_bakta_summary.Forder %>% select(!n.contigs),by = join_by(target_genome == genome))

```


```{r}
genome_summary_df %>% 
  filter(!is.na(prop.contigs.F_order)) %>% pull(all.F_order) %>% table() %>% prop.table()

table(pires1200_bakta_summary.Forder$all.F_order) %>% prop.table()
```








