---
title: "Counting IncF replicons"
author: "Sneha Sundar"
date: "`r Sys.Date()`"
output: html_document
---

Note: Input required for this file is too large for github so this file will not run. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Replicon sequences in each genome were detected using the tool PlasmidFinder (v. 2.1.6) with default parameters (minimum identity 95\% and minimum coverage 60\%). PlasmidFinder detects plasmids in genomes using a pre-defined database of replicon sequences as queries in a BLAST search. The genome was said to contain an F-like replicon if one or more replicons of type IncFIA, IncFIB, IncFIC or IncFII were detected.
Prevalence of F-like replicons was defined as the number of genomes containing an F-like replicon divided by the total number of genomes assessed. 

```{r}
library(tidyverse)
library(here)
```

Load plasmidfinder results
```{r}
plasmidfinder_df <- read.table(here("data/processed/002.find_genes/pires1200/ecoli/plasmidfinder_res/all.tsv"), header = T, sep = "\t",quote = "",)
```

Load genome summary results

```{r}
genome_summary_df <- read.table(here("data/processed/002.find_genes/pires1200_onlyblastcontigs_rmnonF-like_trantotraf/genome_tra_count_summary.tsv"),header = T,sep = '\t', quote = "")

genome_summary_df <- genome_summary_df %>% mutate(functional_category = case_when(
    n.distinct.func.genes < 5 ~ "missing",
    n.distinct.func.genes >=5  & n.distinct.func.genes < 23 ~ "incomplete",
    n.distinct.func.genes   >= 23 ~ "complete",
  ))
  

```


How many replicons per genome do we find? 

```{r}
plasmidfinder_summary <- plasmidfinder_df %>% group_by(genome) %>% summarise(n.replicons = n(), n.incF.replicons=sum(str_detect(Plasmid, "IncF")))

unique(plasmidfinder_df$Plasmid)[startsWith(unique(plasmidfinder_df$Plasmid),"IncF")]
```



```{r}
genome_summary_df <- genome_summary_df %>% left_join(plasmidfinder_summary, join_by(target_genome == genome))
```

```{r}
genome_summary_df <- genome_summary_df %>% mutate(n.replicons = replace_na(n.replicons, 0), 
                             n.incF.replicons = replace_na(n.incF.replicons, 0), has.incF_replicon = n.incF.replicons > 0)
```

## Write replicon count info to overall summary table

```{r}
out_folder <- here("data/processed/002.find_genes/pires1200_onlyblastcontigs_rmnonF-like_trantotraf")


write.table(genome_summary_df, paste0(out_folder, "/","genome_tra_count_summary.tsv"),sep = '\t',quote = F,row.names = F)

```




