---
title: "Genome Quality Checks"
author: "Sneha Sundar"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Script and statistics for Figure S2. 

Analyses of genome quality metrics independently assessed using checkM2 (v.1.1.0) using default settings. 


## Load packages
```{r}

library(tidyverse)
library(here)


plots_folder <- here("data/processed/manuscript_plots/001.sample_distribution")



source(here("code/000.tools/helpers.R"))
source(here("code/000.tools/plot.R"))
```
## Load genome summary results

```{r}
genome_summary_df <- read.table(here("data/processed/002.find_genes/pires1200_onlyblastcontigs_rmnonF-like_trantotraf/genome_tra_count_summary.tsv"),header = T,sep = '\t', quote = "")
```


## Load checkm2 results


```{r}
quality_table_path <- here("data/processed/001.sample_distribution/quality_check/checkm2/quality_report.tsv")

quality_table <- read.table(file = quality_table_path, header = T,sep = "\t")

quality_table
```
Add quality metrics to genome summary table. 

```{r}
genome_summary_df <- right_join(genome_summary_df,quality_table,by=c("target_genome"="Name"))
```

## What is the range of E.coli genomes sizes we have (in bp) ? 

```{r}

range(genome_summary_df$Genome_Size)

```

## Figure S2A: genome completeness and contamination distributions 

```{r}


p <- genome_summary_df %>% ggplot() + 
  geom_jitter(aes(x=Completeness,y=Contamination),alpha=0.5,width = 0.001,height=0.001) + 
 # geom_hline(yintercept=5,linetype=2, linewidth=0.8) +
  xlim(99.98,100.002) + ylim(-0.01,5) +
  ylab("Genome contamination %") + 
  xlab("Genome completeness %") +
  theme_manuscript

p

ggsave(paste0(plots_folder,"/","genome_completeness_contamination.pdf"),p, units = "cm",height = 10, width = 12 , dpi = 600)
  
```


99.5% of genomes have 100% completeness and all genomes have a completeness score >= 99.9%. 

All genomes have less than 3% contamination. 

```{r}
sum(genome_summary_df$Completeness >= 100)

sum(genome_summary_df$Contamination <= 3)
```

## Is genome quality associated with F-like transfer operon functional completness? 

We find a significant difference in mean contamination score across the F-like transfer operon functional completeness categories. Genomes missing F-like transfer operons have statistically lower contamination scores. Note that contamination is in %. Overall, though, contamination scores for all three operon completeness strategies are very low. 


```{r}
kruskal.test(Contamination~functional_category, data=genome_summary_df)
```


```{r}
genome_summary_df %>% group_by(functional_category) %>% summarise(mean(Contamination))
```


```{r}
pairwise.wilcox.test(genome_summary_df$Contamination,genome_summary_df$functional_category,p.adjust.method = "bonferroni")
```

## Figure S2 B,C: genomes completeness and contamination across operon completeness categories

```{r}
p <- genome_summary_df %>%  ggplot(aes(x=functional_category,y=Contamination,color=functional_category)) + 
  geom_violin(scale="count",show.legend=F) +
  geom_jitter(alpha=0.3, width = 0.3) + 
  xlab("F-like transfer operon\nfunctional completeness") + 
  ylab("Genome Contamination (%)") + 
  scale_fill_manual(values=functional_colours)+
  scale_color_manual(values=functional_colours) +
  guides(fill = "none",color="none") + 
  theme_manuscript

p
ggsave(paste0(plots_folder,"/","functional_completness_vs_contamination.pdf"),p, units = "cm",height = 8, width = 8, dpi = 600)
```


```{r}
p <- genome_summary_df %>%  ggplot(aes(x=functional_category,y=Completeness,color=functional_category)) + 
  geom_violin(scale="count",show.legend=F) +
  geom_jitter(alpha=0.3, width=0.3) + 
  xlab("F-like transfer operon\nfunctional completeness") + 
  ylab("Genome Completeness (%)") + 
  scale_fill_manual(values=functional_colours)+
  scale_color_manual(values=functional_colours) +
  guides(fill = "none",color="none") + 
  theme_manuscript
p

ggsave(paste0(plots_folder,"/","functional_completness_vs_genomecompleteness.pdf"),p, units = "cm",height = 8, width = 8, dpi = 600)
```






